Flat-histogram method comparison on 2D Ising Model

### 2020-04-22 ising/ising.tex
making major ising paper edits.

\title{Flat histogram method comparison on 2D Ising Model
}

@@

\section{Introduction}
Flat histogram Monte-Carlo simulation algorithms calculate the thermodynamic
properties of various systems over a range of temperatures.  The first histogram
method used a single canonical Monte Carlo simulation to predict properties for
nearby temperatures~\cite{ferrenberg1988new}. While the method effectively
samples a narrow energy range given limited by a narrow temperature range, it
proves computationally inefficient at sampling large energy ranges.

@@

The WL method can also be implemented as part of a production run which aids in
equilibration~\cite{gross2018massively}. WL is used to generate the weights
resulting in a flat-histogram~\cite{janke2017generalized}. Once a user-defined
minimum $\gamma$ is reached, the simulation switches to a final production run
with fixed weights. WL implemented in this way can ensure ergodicity and
detailed-balance. The converge of the production run is still impacted by the
choice of the minimum $\gamma$.

Because of the uncertainty of convergence for WL, many studies were undertaken
to understand how the modification (or update) factor $\gamma$ impacts the
convergence~\cite{zhou2005understanding,lee2006convergence,
belardinelli2007wang}. The ideal choice of decreasing the modification factor
helps to prevent CPU time from being wasted by continuing to perform
calculations after the error in the density of states is
saturated~\cite{belardinelli2008analysis}. Modification factors that decrease
faster than $1/t$ were found to lead to
non-convergence~\cite{belardinelli2007fast}.  These findings led to the
formation of the $1/t$-WL algorithm.

@@

In this work, we apply the family of weight-based flat histogram Monte
Carlo methods (WL, 1/t-WL, SAMC, SAD) to the 2D Ising model.

\section{Ising Model}
The 2D Ising spin-lattice system is widely used as a testbed when
benchmarking or comparing Monte-Carlo
methods~\cite{ferdinand1969bounded, wang1999transition, trebst2004optimizing, barash2019estimating}. The 2nd order
phase transition behavior and the ability to directly calculate the
exact solution for finite lattices~\cite{beale1996exact, haggkvist2004computation} make the
system sufficiently interesting for such theoretical comparisons. It is
also important to note that direct comparison of the other methods can
be made with WL as its original implementation was done on this
system~\cite{wang2001determining,wang2001efficient}.
We test the convergence of several flat histogram methods
on the periodic 2D square lattice ferromagnetic Ising model with nearest
neighbor interactions~\cite{landau2004new}.
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j - h \sum_i s_i
\end{align}
The $N\times N$ spin system can take on values of $\sigma_i = \pm 1$
for up or down spins respectively. In the absence of a magnetic field ($h =
0$), We can write the Hamiltonian as follows~\cite{onsager1944crystal,
kaufman1949crystal}:
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j
\end{align}
where the sum is over nearest neighbor spin sites. Beale showed that for finite
lattices the Density of States could directly be calculated from the partition
function~\cite{beale1996exact}
\begin{align}
Z = \sum_E g(E) e^{-{\beta E}}
\end{align}
% where $g(E)$ is the multiplicity of the system which is proportional to
% $D(E)$. We can estimate the average deviation from the exact solution
% using the relation~\cite{schneider2017convergence, shakirov2018convergence,
% barash2017control}:
% \begin{align}
% \braket{\epsilon(t)}_E = \frac{1}{N_E - 1} \sum_E \left( \frac{S(E,t) - S_{\text{Beale}(E)}}{S_{\text{Beale}(E)}} \right)
% \end{align}
where $g(E)$ is the multiplicity of the system which is proportional to
$D(E)$. We can estimate the maximum deviation in the canonical specific heat capacity $c_V$ from the exact solution~\cite{schneider2017convergence, shakirov2018convergence,
barash2017control,barash2017gpu}:
\begin{align}
\braket{\epsilon(t)}_E = \max_E \left( c_V(E,S,T,t) - c_V(E,S,T,t)_{\text{Beale}}\right)
\end{align}
% As per the implementation by Wang and
% Landau~\cite{wang2001determining}, the total number of available energy
% states are $N-1$.
Computing the specific heat capacity $c_V$ presents a difficult challenge for
any Monte-Carlo method due to fluctuations in the derivative of the internal energy.  Methods that accurately compute $c_V$ also by extension accurately compute the internal energy.

\section{Flat histogram methods}\label{sec:histogram}
Flat histogram methods determine the density of states $D(E)$ over a broad range
of energies by simulating each energy with equivalent accuracy. Flat histogram
Monte Carlo methods propose randomly chosen ``moves'' which change the state of
the system and must satisfy detailed balance.  Each algorithm differs in how it
determines the probability of accepting a move and in what additional statistics
must be collected in order decide on that probability.

We describe four closely related flat histogram methods which each rely on a
weight function $w(E)$ to determine $D(E)$.  For these algorithms, the
probability of accepting a move is given by
\begin{equation}
	\mathcal{P}(E_\text{old} \rightarrow E_\text{new})
	= \min\left[1,\frac{w(E_\text{old})}{w(E_\text{new})}\right]
\end{equation}
which biases the simulation in favor of energies with low weights. The result of
weights $w(E)$ that are proportional to $D(E)$ is an entirely flat histogram. We
can relate the entropy to the weights in the microcanonical ensemble, since the
entropy is defined as $S(E) \equiv k_B\ln(D(E)) \approx \ln w(E)$.

Each approach uses a random walk in energy space to estimate $D(E)$.  At the
core of each of these approaches is to continuously update the weights at each
step of the simulation
\begin{equation}
	\ln{w_{t+1}(E)}=\ln{w_{t}(E)}
	+\gamma_t
\end{equation}
where $t$ is number of the current move, $\gamma(t)$ is a move-dependent update
factor, and $E$ is the current energy.  This update causes the random walk to
avoid energies that have been frequently sampled, leading to a rapid exploration
of energy space. The four methods differ primarily in how they schedule the
decreasing of $\gamma_t$.

The Wang-Landau algorithm~\cite{wang2001efficient,wang2001determining,
landau2014guide} explores energy space by initially setting $\gamma=1$, and then
decreases $\gamma$ in discrete stages. An energy range of interest must be
specified~\cite{wang2001efficient, schulz2003avoiding, yan2003fast},
which can require multiple simulations if unknown.

The number (``counts'') of moves ending at each energy are stored in a
histogram.  For a sufficiently flat energy histogram (specified by the user to
typically be 0.8), $\gamma$ is decreased by a specified factor of $\frac12$ and
the histogram is reset to zero. The entire process is repeated until $\gamma$
reaches a desired cutoff.

@@

\jpsays{consider adding in minGamma WL here or before 1/t-WL method.
An alternative method of implementing the WL algorithm is to first implement a “production” run~\cite{}. The WL method updates the weights w(E) until some Gamma min is reached. At this point, the weights are fixed and the method proceeds.}

Another weight-based flat histogram method is
the Stochastic Approximation Monte Carlo (SAMC) algorithm. SAMC has a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} =\frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.

SAMC offers extreme simplicity and proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
liang2007stochastic}.  Unlike WL methods, the energy range need not be
known \emph{a priori.}  The convergenc time depends only on the choice of
parameter $t_0$.
The parameter $t_0$ can unfortunately be difficult to chose in advance
for arbitrary systems.
Liang \emph{et al.} give a rule of thumb in
which $t_0$ is chosen in the range from $2N_S$ to $100N_S$ where $N_S$
is the number of energy bins~\cite{liang2007stochastic}.  Schneider
\emph{et al.} found that for the Ising model this heuristic is helpful
for small spin systems, but that larger systems require an even higher
$t_0$ value~\cite{schneider2017convergence}. Our work also confirms this.

>>>>

\title{Flat-histogram method comparison on 2D Ising Model
}

@@

\section{Introduction}
Flat-histogram Monte-Carlo simulation algorithms calculate the thermodynamic
properties of various systems over a range of temperatures.  The first histogram
method used a single canonical Monte Carlo simulation to predict properties for
nearby temperatures~\cite{ferrenberg1988new}. While the method effectively
samples a narrow energy range given limited by a narrow temperature range, it
proves computationally inefficient at sampling large energy ranges.

@@

Because of the uncertainty of convergence for WL, many studies were undertaken
to understand how the modification (or update) factor $\gamma$ impacts the
convergence~\cite{zhou2005understanding,lee2006convergence,
belardinelli2007wang}. The ideal choice of decreasing the modification factor
helps to prevent CPU time from being wasted by continuing to perform
calculations after the error in the density of states is
saturated~\cite{belardinelli2008analysis}. Modification factors that decrease
faster than $1/t$ were found to lead to
non-convergence~\cite{belardinelli2007fast}.  These findings led to the
formation of the $1/t$-WL algorithm and also sparked some debate as to whether
the simulation should be split into multiple parts, called ``production run" WL, in order to preserve ergodicity and
detailed balance~\cite{jayasri2005wang, mukhopadhyay2008monte}.

@@

In this work, we apply the family of weight-based flat-histogram Monte
Carlo methods (WL, 1/t-WL, SAMC, SAD) to the 2D Ising model.

\section{Ising Model}
The 2D Ising spin-lattice system is widely used as a testbed when
benchmarking or comparing Monte-Carlo
methods~\cite{ferdinand1969bounded, wang1999transition, trebst2004optimizing, barash2019estimating}. The 2nd order
phase transition behavior and the ability to directly calculate the
exact solution for finite lattices~\cite{beale1996exact, haggkvist2004computation} make the
system sufficiently interesting for such theoretical comparisons. It is
also important to note that direct comparison of the other methods can
be made with WL as its original implementation was done on this
system~\cite{wang2001determining,wang2001efficient}.
We test the convergence of several flat-histogram methods
on the periodic 2D square lattice ferromagnetic Ising model with nearest
neighbor interactions~\cite{landau2004new}.
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j - h \sum_i s_i
\end{align}
The $N\times N$ spin system can take on values of $\sigma_i = \pm 1$
for up or down spins respectively. In the absence of a magnetic field ($h =
0$), We can write the Hamiltonian as follows~\cite{onsager1944crystal,
kaufman1949crystal}:
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j
\end{align}
where the sum is over nearest neighbor spin sites. Beale showed that for finite
lattices the Density of States could directly be calculated from the partition
function~\cite{beale1996exact}
\begin{align}
Z = \sum_E g(E) e^{-{\beta E}}
\end{align}
% where $g(E)$ is the multiplicity of the system which is proportional to
% $D(E)$. We can estimate the average deviation from the exact solution
% using the relation~\cite{schneider2017convergence, shakirov2018convergence,
% barash2017control}:
% \begin{align}
% \braket{\epsilon(t)}_E = \frac{1}{N_E - 1} \sum_E \left( \frac{S(E,t) - S_{\text{Beale}(E)}}{S_{\text{Beale}(E)}} \right)
% \end{align}
where $g(E)$ is the multiplicity of the system which is proportional to
$D(E)$. We can estimate the maximum deviation in the canonical specific heat capacity $c_V$ from the exact solution~\cite{schneider2017convergence, shakirov2018convergence,
barash2017control,barash2017gpu}:
\begin{align}
\braket{\epsilon(t)}_E = \max_E \left( c_V(E,S,T,t) - c_V(E,S,T,t)_{\text{Beale}}\right)
\end{align}
% As per the implementation by Wang and
% Landau~\cite{wang2001determining}, the total number of available energy
% states are $N-1$.
Computing the specific heat capacity $c_V$ presents a difficult challenge for
any Monte-Carlo method due to fluctuations in the derivative of the internal energy.  Methods that accurately compute $c_V$ also by extension accurately compute the internal energy.

\section{Flat-histogram methods}\label{sec:histogram}
Flat-histogram methods determine the density of states $D(E)$ over a broad range
of energies by simulating each energy with equivalent accuracy. Flat-histogram
Monte Carlo methods propose randomly chosen ``moves'' which change the state of
the system and must satisfy detailed balance.  Each algorithm differs in how it
determines the probability of accepting a move and in what additional statistics
must be collected in order decide on that probability.

We describe several closely related flat-histogram methods which each rely on a
weight function $w(E)$ to determine $D(E)$.  For these algorithms, the
probability of accepting a move is given by
\begin{equation}
	\mathcal{P}(E_\text{old} \rightarrow E_\text{new})
	= \min\left[1,\frac{w(E_\text{old})}{w(E_\text{new})}\right]
\end{equation}
which biases the simulation in favor of energies with low weights. The result of
weights $w(E)$ that are proportional to $D(E)$ is an entirely flat-histogram. We
can relate the entropy to the weights in the microcanonical ensemble, since the
entropy is defined as $S(E) \equiv k_B\ln(D(E)) \approx \ln w(E)$.

Flat-histogram methods employ a random walk in energy space to estimate $D(E)$.  Each method operates by continuously updating the weights at each
step of the simulation
\begin{equation}
	\ln{w_{t+1}(E)}=\ln{w_{t}(E)}
	+\gamma_t
\end{equation}
where $t$ is number of the current move, $\gamma(t)$ is a move-dependent update
factor, and $E$ is the current energy.  This update causes the random walk to
avoid frequently sampling the same energies, leading to a rapid exploration
of energy space. Flat-histogram methods differ primarily in how they schedule
the decrease of $\gamma_t$.

The Wang-Landau algorithm~\cite{wang2001efficient,wang2001determining,
landau2014guide} explores energy space by setting $\gamma_{t=0}^{\text{WL}}=1$,and then decreases $\gamma^{\text{WL}}$ in prescribed stages. An energy range of
interest must be specified~\cite{wang2001efficient, schulz2003avoiding,yan2003fast}, which often requires multiple simulations if unknown.

The number (``counts'') of moves ending at each energy are stored in a
histogram.  For a sufficiently flat energy histogram (typically user-specified to be 0.8), $\gamma^{\text{WL}}$ is decreased by a specified factor of $\frac12$ and the histogram is reset to zero. The entire process is repeateduntil $\gamma^{\text{WL}}$ reaches a desired cutoff.

@@

The WL method can also be implemented as part of a production run which aids in
equilibration~\cite{gross2018massively}. WL is used to generate the weights
resulting in a flat-histogram~\cite{janke2017generalized}. Once a user-defined
minimum $\gamma$ is reached, the simulation switches to a final production run
with fixed weights. WL implemented in this way can ensure ergodicity and
detailed-balance; however, the convergence of simulation is still impacted by
the choice of the minimum $\gamma$.

Another weight-based flat-histogram method is
the Stochastic Approximation Monte Carlo (SAMC) algorithm. SAMC has a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} =\frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.

SAMC offers extreme simplicity and proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
liang2007stochastic}.  Unlike WL methods, the energy range need not be
known \emph{a priori.}  The convergenc time depends only on the choice of
parameter $t_0$.
The parameter $t_0$ can unfortunately be difficult to chose in advance
for arbitrary systems.
Liang \emph{et al.} give a rule of thumb in
which $t_0$ is chosen in the range from $2N_S$ to $100N_S$ where $N_S$
is the number of energy bins~\cite{liang2007stochastic}.  Schneider
\emph{et al.} found and we confirm that for the Ising model this heuristic is helpful for small spin systems, but that larger systems require an even higher
$t_0$ value~\cite{schneider2017convergence}.
